---
title: "Model examples"
date: last-modified
toc: true
format: 
  html:
    output-dir: docs 
---

This is an example of using three different methods to analyze data with the following characteristics.

* Multiple observations per individual
* A discrete independent variable

# Read packages

```{r}
library(tidyverse) # Modern R
library(emmeans)   # Pair-wise comparisons
library(lme4)      # Random effect models
library(lmerTest)  # ANOVA table for random effect model
```

# Prepare demonstration data

```{r}
dset = as_tibble(iris)
dset = dset |> 
  mutate(Sepal = Sepal.Length * Sepal.Width,
         Petal = Petal.Length * Petal.Width) |> 
  select(Species, Sepal, Petal) |> 
  group_by(Species) |> 
  slice_sample(n = 20) |> 
  mutate(n = 1:n()) |> 
  ungroup()
```

This data is a modified version from the iris dataset in R.
I calculated the area of the Sepal and Petal for each Species (*I. setosa*, *I. versicolor*, and *I. virginica*),
so that it appears similar to your fish gill-branch area data.

```{r}
ggplot(dset) + 
  geom_point(aes(x = Sepal, y = Petal, color = Species)) +
  labs(title = "The area of the sepal and petal for each Iris species.")
```

# Traditional one-way ANOVA

In a simple one-way ANOVA, we first sum the area of the sepal and the petal.

```{r}
d1 = dset |> 
  mutate(Area = Sepal + Petal)
```

```{r}
ggplot(d1) + 
  geom_point(aes(x = Species, y = Area, color = Species),
             position = position_jitter(0.1)) +
  labs(title = "Is the mean area different among treatment effect?")
```

Then we test the null hypothesis that the mean value of the Area is similar across treatment effect (Species).



```{r}
m1 = lm(Area ~ Species, data = d1)
summary.aov(m1)
```

At a significance level of 0.05, the results are significant.
Next we could use a TukeyHSD test to test for post-hoc pair-wise differences in the mean area.

```{r}
emmeans(m1, specs = pairwise ~ Species, type = "Tukey")
```

If we look a the `$contrast` output, we can see that the difference are significant among all pair-wise comparisons.

# Multivariate ANOVA (MANOVA)

My preferred method for this type of data, where there are multiple dependent variables for each sample, would be a multivariate ANOVA.
In this case, Sepal and Petal are multiple observations for each sample (i.e., dependent variable) and the Species is the independent variable.

```{r}
ggplot(d1) + 
  geom_point(aes(x = Sepal, y = Petal, color = Species)) +
  labs(title = "Is the mean area of the Sepal and Petal different among treatment effect?")
```

Here, we test the hypothesis that the mean area of the Sepal and Petal are similar across treatment effect (Species).

```{r}
m2 = lm(cbind(Sepal, Petal) ~ Species, data = d1)
manova_result = manova(m2)
summary(manova_result, test = "Pillai")
```

According to the Pillai statistic, the null hypothesis can be rejected at the 0.05 significance level.

```{r}
emmeans(manova_result, specs = pairwise ~ Species, type = "Tukey")
```
The TukeyHSD test also indicates significant difference in pair-wise comparisons.

# One-way ANOVA with a random effect

Random effect models are very tricky.
Please read Ben Bolker's FAQ on GLMMs, which deals with random effect models.
Especially read the discussion regardin "Should I treat factor xxx as fixed or random?" 

URL: <https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html>

In this case the fixed effect is Species and the random effect is Organ.
In your fish data, the fixed effect would be the *treatment* and the random effect would be *gill arch #*.

```{r}
dset2 = 
  dset |> 
  pivot_longer(cols = c(Sepal, Petal),
               names_to = "Organ",
               values_to = "Area") |> 
  mutate(Organ = factor(Organ),
         Species = factor(Species))
```

The null hypothesis for the fixed effect is similar to that of the simple ANOVA.

```{r}
m3 = lmer(Area ~ Species +(1 | Organ), data = dset2)
summary(m3)
```

Note there is output for randome effect, fixed effect, and the correlation among fixed effects.
The P-values for the random effects model is provided below and indicates that null hypothesis can be rejected.

```{r}
anova(m3)
```

The TukeyHSD test also indicates significant difference in pair-wise comparisons.


```{r}
emmeans(m3, specs = pairwise ~ Species, type = "Tukey")
```
